---
title: "Identifying_impacts_extreme_weather"
author: "Marie Tolteca"
date: 11/10/2025
format: html
---
#### Import Libraries

```{r, message = FALSE}
library(terra) # raster handling
library(tidyverse)
library(tmap) # map making
library(kableExtra) # table formatting
library(spData) # spatial data
library(spDataLarge) # spatial data
library(geodata) # spatial data
library(here)
library(sf)
library(raster)
library(stars)
library(patchwork)
```

#### Importing in Data
```{r, message=FALSE, warning=FALSE, output = FALSE}
texas <- st_read(here::here('data','ACS_2019_5YR_TRACT_48_TEXAS.gdb'),
                 layer ='ACS_2019_5YR_TRACT_48_TEXAS' ,
                 # Adding to not show output message
                 quiet = TRUE)


homes <- st_read(here('data','gis_osm_buildings_a_free_1.gpkg'), 
                 query = "SELECT * FROM gis_osm_buildings_a_free_1 WHERE (type IS NULL AND name IS NULL)",
                # Adding to not show output message
                 quiet = TRUE)


roads <- st_read(here('data','gis_osm_roads_free_1.gpkg'),
                 query = "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'",
                # Adding to not show output message
                 quiet = TRUE)

# Raster Data - 
vnp_05 <- rast(here("data", "VNP46A1", "VNP46A1.A2021038.h08v05.001.2021039064328.tif"))
vnp_06 <- rast(here("data", "VNP46A1", "VNP46A1.A2021038.h08v06.001.2021039064329.tif"))
vnp_05_2 <- rast(here("data", "VNP46A1", "VNP46A1.A2021047.h08v05.001.2021048091106.tif"))
vnp_06_2 <- rast(here("data", "VNP46A1", "VNP46A1.A2021047.h08v06.001.2021048091105.tif"))

# Using Stars for before and after
vnp_051 <- read_stars(here("data", "VNP46A1", "VNP46A1.A2021038.h08v05.001.2021039064328.tif"))
vnp_061 <- read_stars(here("data", "VNP46A1", "VNP46A1.A2021038.h08v06.001.2021039064329.tif"))
vnp_05_21 <- read_stars(here("data", "VNP46A1", "VNP46A1.A2021047.h08v05.001.2021048091106.tif"))
vnp_06_21 <- read_stars(here("data", "VNP46A1", "VNP46A1.A2021047.h08v06.001.2021048091105.tif"))

```

#### Creating Raster Objects
```{r, message=FALSE, warning=FALSE}
# Checking Extents:
if(st_crs(vnp_05) == st_crs(vnp_06)){
  print("Extents match :)")
}  else{
  warning("extents do not match :(")
}

# Stars
vnp_feb7<- st_mosaic(vnp_051,vnp_061)
vnp_feb16 <- st_mosaic(vnp_05_21, vnp_06_21)

#Cropping extent
extent <- st_bbox(c(xmin = -96.5, # xmin
                  ymin = 29, # ymin
                  xmax = -94.5, # xmax
                  ymax = 30.5),
                  crs = st_crs(vnp_feb7))
# Arranging plots- side by side
par(mfrow = c(1, 2))

#Cropped VNP
vnp_7_crop <- st_crop(vnp_feb7, extent)

vnp_16_crop <- st_crop(vnp_feb16, extent)

vnp_07 <- plot(vnp_7_crop, main = "Before Storm", key.pos = NULL) 
vnp_16 <- plot(vnp_16_crop, main = "After Storm", key.pos = NULL)

# Saving Image
ggsave("vnp_07_before.png", plot = vnp_07, width = 8, height = 6, units = "in", dpi = 300)
ggsave("vnp_16_after.png", plot = vnp_16, width = 8, height = 6, units = "in", dpi = 300)
```
#### Using Terra Packages 
```{r}
# Combining Rasters: Using xmin and xmax
vnp_feb07 <- merge(vnp_05, vnp_06)
vnp_feb16 <- merge(vnp_05_2, vnp_06_2)


# Checking if bounding box are the same
st_bbox(vnp_feb07)
st_bbox(vnp_feb16)

# Computing the Difference
vnp_change <- (vnp_feb07) - (vnp_feb16) # Using Terra
```



#### Reclassify the difference raster, assuming that any location that experienced a *drop of more than 200 nW cm-2sr-1* experienced a blackout
```{r, message=FALSE, warning=FALSE}
# Define the reclassification matrix 
rcl_vnp <- matrix(c(-Inf, 200, NA,   # values -Inf to 200 = NA
                    200, Inf, 1),    # values 200 to Inf = 1
                  ncol = 3, byrow = TRUE)

# Apply the matrix using terra::classify
reclassified <- terra::classify(vnp_change, rcl = rcl_vnp)
# Assinging map to a variable
reclassified_map <- tm_shape(reclassified)+
  tm_raster(palette="viridis", midpoint=0)+
  tm_title("Reclassified Difference VNP")
# Viewing map
reclassified_map

# Saving Map
tmap_save(reclassified_map,filename = here("images", "reclassified_map.png"),width = 1200, height = 800, dpi = 300)

```

#### Vectorize the blackout mask
hint: use st_as_sf() to convert from a raster to a vector and fix any invalid geometries with st_make_valid()
```{r}
blackout_vector <- st_as_stars(reclassified)%>% 
  st_as_sf() %>% 
  st_make_valid()
```
#### Crop(spatially subset) the blackout mask to the Houston area as defined by the following coordinates:
(-96.5, 29), (-96.5, 30.5), (-94.5, 30.5), (-94.5, 29)
```{r, warning=FALSE, message=FALSE}
# Define bounding box coordinates
blackout_houston_bounds <- c(xmin = -96.5, xmax = -94.5,
                             ymin = 29, ymax = 30.5)

# Creating bbox and assign CRS
houston_bbox <- st_bbox(blackout_houston_bounds, crs = st_crs(vnp_feb07))

# Convert bbox to an sf polygon
houston_bbox_sf <- st_as_sfc(houston_bbox)

# Crop the vectorized blackout map to Houston area
blackout_houston <- st_intersection(blackout_vector, houston_bbox_sf)

# Re-project the cropped blackout dataset to *EPSG:3083* (NAD83 / Texas Centric Albers Equal Area)?
blackout_houston_proj <- st_transform(blackout_houston, crs = 3083)

# Checking CRS
#st_crs(blackout_houston_proj)

blackout_houston <- tm_shape(blackout_houston_proj) +
    # Add OpenStreetMap as the basemap
  tm_basemap(server = "OpenStreetMap") +
  tm_polygons()+
  tm_layout(main.title = "Blackout Map of Houston, Texas",
            main.title.size = 1.5)+
  tm_add_legend(
    type = "fill",
    labels = "Blackout Areas",
    col = "gray30",
    title = "Locations"
  ) +
  tm_compass(position = "left", "bottom")+
  tm_scale_bar(position = "left", "bottom")
# Viewing
blackout_houston

# Saving Map
tmap_save(blackout_houston,filename = here("images", "blackout_houston_map.png"),width = 1200, height = 800, dpi = 300)
```

#### Exclude highways from the cropped blackout mask and Identify areas within 200m of all highways - hint: you may need to use st_union
```{r, message=FALSE}
# Checking Extents:
if(st_crs(blackout_houston_proj) == st_crs(roads)){
  print("Extents match :)")
}  else{
  warning("extents do not match :(")
}
#Checking Units
st_crs(roads, parameters = TRUE)$units_gdal # Degrees

# Transform
highway <- st_transform(roads, st_crs(blackout_houston_proj))

#Checking Units
st_crs(highway, parameters = TRUE)$units_gdal # Metres

# Confirming Extents:
if(st_crs(blackout_houston_proj) == st_crs(highway)){
  print("Extents match :)")
}  else{
  warning("extents do not match :(")
}

# Creating buffer
highway_200 <- st_buffer(highway, dist = 200) %>% 
  st_union()

# Difference in blackout houston project and highway buffer 200 m
highway_blackouts <- st_difference(blackout_houston_proj, highway_200)


# MAP
highway200_buff <- tm_basemap("OpenStreetMap") +
  tm_shape(highway_blackouts) +
  tm_polygons(col = "gray30", border.col = NA, alpha = 0.6) +
  tm_shape(highway_200) +
  tm_borders(col = "deeppink", lwd = 1) +
  tm_layout(main.title = "Houston Blackouts Highways(Buffer of 200m)",
            main.title.size = 1.5)+
  tm_add_legend(
    type = "fill",
    labels = c("Blackout Area (Excluding Highways)", "Highways (200 m Buffer)"),
    col = c("gray30", "deeppink"),
    title = "Legend"
  )+
  tm_compass(position = "left", "bottom")+
  tm_scale_bar(position = "left", "bottom")
# Viewing
highway200_buff

# Saving Map
tmap_save(highway200_buff,filename = here("images", "highway200_buff.png"),width = 1200, height = 800, dpi = 300)

```

#### Identify the number of homes likely impacted by blackouts
- identify homes that overlap with areas that experienced blackouts
```{r, message=FALSE}
# Transform homes to match CRS
buildings <- st_transform(homes, st_crs(highway_blackouts))

# Verify
if (st_crs(highway_blackouts) == st_crs(buildings)) {
  print("CRS match :)")
} else {
  warning("CRS do not match ️:(")
}

# homes that overlap with blackout areas
building_overlaps <- st_intersects(buildings, highway_blackouts)


if (st_crs(highway_blackouts) == st_crs(buildings)) {
  print("CRS match :)")
} else {
  warning("CRS do not match ️:(")
}


# Boolean TRUE/FALSE for homes impacted
buildings$impacted <- lengths(building_overlaps) > 0

# Subset homes that were affected
blackout_buildings <- buildings[buildings$impacted, ]

# How many homes are impacted
impacted_count <- sum(buildings$impacted)
print(paste("Number of homes likely impacted by blackouts:", impacted_count))

# Switching polygons to points
building_points <- st_centroid(buildings)

# Switching polygons to points
blackout_buildings_points <- st_centroid(blackout_buildings)

# MAP
building_impacted <- tm_shape(building_points) +
  # Add OpenStreetMap as the basemap
  tm_basemap(server = "OpenStreetMap") + 
  tm_dots(col = "grey30", size = 0.05, alpha = 0.5,
          title = "Homes (Not Impacted)") +
tm_shape(blackout_buildings_points) +
  tm_dots(col = "darkorchid", size = 0.05, alpha = 0.8,
          title = "Homes Impacted by Blackouts") +
  tm_layout(main.title = "Homes Impacted by Blackout",
            main.title.size = 1.2)+
tm_add_legend(type = "symbol",
              labels = c("Not Impacted", "Impacted by Blackouts"),
              col = c("grey30", "darkorchid"),
              title = "Homes")+
  tm_compass(position = "left", "bottom")+
  tm_scale_bar(position = "left", "bottom")

# Viewing
building_impacted

# Saving Map
tmap_save(building_impacted,filename = here("images", "building_impacted.png"),width = 1200, height = 800, dpi = 300)
```


#### Socioeconomic Houston Texas 
```{r, output = FALSE, echo=FALSE}
# Viewing what is in the geodatabase
st_layers(here::here('data','ACS_2019_5YR_TRACT_48_TEXAS.gdb'))

# Read in geometry (tract boundaries)
geometry_socio <- st_read(here('data','ACS_2019_5YR_TRACT_48_TEXAS.gdb'),
                          layer = 'ACS_2019_5YR_TRACT_48_TEXAS')

# Read in income data
income <- st_read(here('data','ACS_2019_5YR_TRACT_48_TEXAS.gdb'),
                  layer = 'X19_INCOME')

# Join geometry with income attributes
socioeconomic <- geometry_socio %>% 
  left_join(income, by = c("GEOID_Data" = "GEOID"))

# Check for missing values
colSums(is.na(socioeconomic))

# Reproject socioeconomic data to match blackout CRS
socioeconomic_proj <- st_transform(socioeconomic, st_crs(highway_blackouts))

# Identify tracts that intersect blackout areas
tracts_blackout <- st_intersects(socioeconomic_proj, highway_blackouts)

# Create a logical indicator for blackout exposure
socioeconomic_proj$blackout <- lengths(tracts_blackout) > 0

# Make a human-readable label for plotting
socioeconomic_proj <- socioeconomic_proj %>%
  # Creating new column
  mutate(blackout_status = ifelse(blackout, "Affected by Blackout", "Not Affected"))
```
#### Visual of sociaeconomic median
```{r, warning=FALSE}
# Boxplot comparing income distributions ---
median_income <- ggplot(socioeconomic_proj, aes(x = blackout_status, y = B19013e1, fill = blackout_status)) +
  geom_boxplot(width = 0.6, outlier.alpha = 0.4) +
  scale_fill_manual(values = c("Affected" = "grey30", 
                               "Not Affected" = "hotpink")) +
  labs(
    title = "Median Household Income for Census Tracts in Houston",
    subtitle = "Comparing tracts affected and not affected by blackouts",
    x = "",
    y = "Median Household Income (USD)"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.text.x = element_text(face = "bold"),
    legend.position = "none"
  )

# Viewing
median_income

# Saving Map
ggsave(filename = here("images", "median_income.png"), plot =median_income , width = 8, height = 6, dpi = 300)
```

#### Income Summary Table
```{r}
income_summary <- socioeconomic_proj %>%
  # Drop geometry for speed
  st_drop_geometry() %>% 
  # Group by blackout
  group_by(blackout) %>%
  # Grab median incomes- using metadata
  summarize(
    mean_income = mean(as.numeric(B19013e1), na.rm = TRUE),
    median_income = median(as.numeric(B19013e1), na.rm = TRUE),
    sd_income = sd(as.numeric(B19013e1), na.rm = TRUE),
    n_tracts = n()
  ) %>%
  # Creating new column
  mutate(blackout = ifelse(blackout, "Blackout", "No Blackout"))

# Kable Extra
income_summary %>%
  kbl(
    caption = "Summary Statistics: Median Household Income by Blackout Status",
    col.names = c("Blackout Status", "Mean Income ($)", "Median Income ($)", "SD", "Number of Tracts"),
    digits = 0
  ) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(2:4, width = "3cm") %>%
  add_header_above(c(" " = 1, "Income Summary" = 3, " " = 1))


```





